---
title: "R Demo"
output: html_notebook
---

### load extra library, import data
```{r}
library(tidyverse)

data <- read_csv("~/git/BIS180L_web/data/Tomato.csv")
```

### summaries
```{r}
#get a quick summary
summary(data)

#look at the first few lines
head(data)
```

### a histogram

```{r}
ggplot(data,aes(x=hyp)) + geom_histogram()
```



### histograms of hyp, split by treatment

```{r}
ggplot(data,aes(x=hyp)) + 
  geom_histogram() +
  facet_grid(. ~ trt)
```

### histograms of hyp, split by species

```{r}
ggplot(data,aes(x=hyp)) + 
  geom_histogram() +
  facet_grid(. ~ species)
```


### histograms of hyp, split by treatment and species
```{r}
ggplot(data,aes(x=hyp)) + 
  geom_histogram() +
  facet_grid(trt ~ species)
```

### look at individual data points
```{r}
qplot(x=1,y=hyp,data=data, xlab="")
```

### spread it out some
```{r}
qplot(x=1,y=hyp,data=data,geom="jitter", xlab="")
```

### split it by species
```{r}
qplot(x=species,y=hyp,data=data,geom="jitter")
```


### indicate treatment by color
```{r}
qplot(x=species,y=hyp,data=data,color=trt,geom="jitter")
```

### indicate measure by plotting symbol
```{r}
qplot(x=species,y=hyp,data=data,colour=trt,shape=who,geom="jitter")
```


### scatter plots 
```{r}
#how about the relationship between two internodes?
qplot(data$int1,data$int2)
```

```{r}
#between all internode data?
pairs(data[,c("int1","int2","int3","int4")])
```

```{r}
cor(data[,c("int1","int2","int3","int4")],use="complete.obs")
```



## some simple statistics:

```{r}
#mean
mean(data$hyp)

#median
median(data$hyp)

#variance
var(data$hyp)

#standard dev
sd(data$hyp)
```

#create our own sem function

```{r}
sem <- function(x) {
	sd(x,na.rm=T)/sqrt(length(na.omit(x)))}
```


### calculate means for each species and treatment separately

```{r}
data %>% group_by(species,trt) %>%
  summarize(mean=mean(hyp,na.rm = T))
```


## calculate mean and s.e.m. for each species and treatment

```{r}
data %>% group_by(species,trt) %>%
  summarize(mean=mean(hyp,na.rm=TRUE),sem=sem(hyp))
```

### simple statistical tests

```{r}
t.test(data$hyp~data$trt)
```

### all species t tests
```{r}
pairwise.t.test(data$hyp,data$species)

```

### calculate ANOVA, put results in an object and summarize
```{r}
aov1 <- aov(data$hyp~data$species)
summary(aov1)
```

### more anova
```{r}
#two way
aov2 <- aov(hyp~species+trt,data=data)
summary(aov2)

#two-way with interaction
aov3 <- aov(hyp~species*trt,data=data)
summary(aov3)
```


### linear regression

```{r}
#linear regressions have nicer outputs
#calculate linear regression, put results in an object and summarize
lm1 <- lm(hyp~species,data=data)
summary(lm1)

#two way
lm2 <- lm(hyp~species+trt,data=data)
summary(lm2)

#two way with interaction effect
lm3 <- lm(hyp~species*trt,data=data)
summary(lm3)
```

